Letter from ANU

### 2025-07-25 cover_letter/template.qmd
update letter

This paper builds on our previous work, "A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol" (JCGS), which provided empirical evidence for the effectiveness of visual evaluation of residual plots through crowd-sourced lineup experiments. While that study highlighted the cost and effort of human evaluations, this work advances automation through computer vision. We introduce a model trained to predict a custom distance measure that quantifies the discrepancy between residuals and theoretically "good" distributions. Building on these predictions, we construct a statistical testing framework aligned with the original lineup protocol, ensuring a valid visual test. When evaluated on the earlier experiment data, the model performs slightly worse than humans but surpasses conventional statistical tests due to its reduced sensitivity to minor deviations from model assumptions. We further demonstrate that the model, like human observers, relies on visual patterns and shapes, making it a practical tool for reducing manual effort in residual plot diagnostics. This approach should be of particular interest to readers concerned with model diagnostics and the automation of visual data analysis.

>>>>

This paper builds on our previous work, "A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol" (JCGS), which provided empirical evidence for the effectiveness of visual evaluation of residual plots through crowd-sourced lineup experiments. While that study highlighted the cost and effort of human evaluations, this work advances automation of residual plot diagnostics through an application of a computer vision model. We specifically trained a model to predict a custom distance measure that quantifies the discrepancy between residuals and theoretically "good" distributions. Building on these predictions, we construct a statistical testing framework aligned with the original lineup protocol, ensuring a valid visual test. When evaluated on the earlier experimenal data, the model performs slightly worse than humans but surpasses conventional statistical tests due to its reduced sensitivity to minor deviations from model assumptions. We further demonstrate that the model, like human observers, relies on visual patterns and shapes, making it a practical tool for reducing manual effort in residual plot diagnostics. This approach would be of particular interest to readers concerned with model diagnostics and the automation of visual data analysis.
